Project Manager's Progress Report

### 2016-01-16 pnsac/newsletter/article_Feature1.tex
Adding files for April 2015 newsletter.

\textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
  Sqn, RCAF.}

@@

\begin{figure}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{jim_riddoch.eps}
   %caption of the figure 
   \caption*{\small \em Jim Riddoch.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:jim_riddoch.eps}
\end{figure}

In 1947, while Prime Minister Mackenzie King was in London, his
Minister of Foreign Affairs (and chosen suc-cessor), Louis St-Laurent,
consented to Canadian participation on the United Nations Temporary
Commission of Korea (UNTOK). On his return to Canada, King furiously
confronted St-Laurent. He wanted no part---even at the fringes—of
Korean affairs, which he considered dangerous.

@@

Mackenzie King died on July 22, 1950. As he lay in state in the
Parliament Buildings, six North Star aircraft of No. 426 Squadron flew
over, saluting his passing as they were en route to Tacoma,
Washington, to commence a Royal Canadian Air Force transoceanic
airlift in sup-port of United Nations forces fighting in Korea.

It is not known if the former prime minister was spin-ning in his
coffin.

@@

The North Star was basically a Douglas DC-4, but with Rolls-Royce
Merlin in-line engines instead of Pratt-andWhitney radials. It was a
dependable aircraft, but generations of passengers and crew would
remember the tremendous noise and vibration in flight. The full story
of this is related by Larry Milberry in his 1982 book The Canadair
North Star.

@@

While the RCAF did not participate in the July 1948 to May 1949 Berlin
Airlift, a few individual aircrew did while on exchange duties with
the Royal Air Force and United States Air Force. As it happened, the
airlift seriously degraded the USAF Military Air Transport Service
(MATS) after flour and coal dust penetrated the seams of every
participating aircraft. When it was next marshalled to meet the Korean
crisis, MATS needed all the help it could get—and that was forthcoming
when the Canadian government committed No. 426 Sqdn.—its only
long-range transport unit—to the Korean campaign.

In recounting the RCAF’s transpacific operational story, one must pay
tribute to Laurence Motiuk, author of two monumental volumes, namely
Thunderbirds at War: Diary of a Bomber Squadron, and Thunderbirds for
Peace: Diary of a Transport Squadron, which set out the history of
No. 426 Sqdn. The latter work was particularly difficult to write,
given that in the 1950s, RCAF units often submitted incomplete
historical narratives and in some instances neglected to prepare any
at all.

Motiuk’s account of the postwar squadron was drawn from numerous
unofficial records, including log books as well as the sketchy unit
reports.

@@

Deployment to McChord Air Force Base, Tacoma, Washington, began on the
25th. Six North Stars, 12 crews and 185 ground personnel were
dispatched. Twothirds of the aircrew were veterans of the Second
World War; the remainder were men who had joined the RCAF after
1947. Three of the aircraft left Tacoma on the 27th, arriving at
Haneda in Tokyo two days later. Operation Hawk was under way.

@@

The North Stars followed a challenging route. From McChord to
Elmendorf Air Force Base in Alaska was 1,490 miles. This was followed
by a 1,537-mile run to Shemya Air Force Base in the Aleutian
Islands. Next came the 2,104mile flight to Haneda in Tokyo, with the
aircraft making landfall at Matsushima (Honshu, Japan).

@@

Between July 28 and Dec. 31, 1950, the squadron dispatched 123
aircraft across the Pacific. In the whole of 1951 it was 193 missions,
but in 1952 it dropped to 133. Not only were the Thunderbirds
returning to more traditional duties in Canada, they were also
assuming the job of providing transport, including personnel
rotation, to Canada’s growing land and air presence in Europe. In
1953—the last full year of Hawk—it was 85 missions.

The final Hawk mission departed McChord on May 31, 1954, and ended at
the squadron’s home base at Dorval on June 9, 1954. As it wound up, an
RCAF dispatch reported that No. 426 Sqdn. had flown 599 round
trips—four of them directly to Korea itself—logged 34,000 flying
hours, carried 13,000 personnel, and airlifted 3,500 tons of
freight. The report erred on two points. Six flights (not four) had
gone directly to Korea. The other discrepancy was that there were
only 584 missions; for some unknown reason the numbered listing of
flights had jumped from “439” on Nov. 26, 1952, to “455” on Nov. 28,
1952; the inaccurate 1954 press release continues to be quoted, ad
nauseam, to this day.

@@

On April 19, 1951, a North Star piloted by Flight Lieutenant J.A. Watt
was on a route-training flight to an unfamiliar field—Ashiya—located
between Osaka and Kobe, Japan. He was cleared by Ashiya tower to
descend from 4,000 to 3,000 feet. At 3,400 feet the aircraft hit trees
atop a hill that was not even indicated on the airfield map. No one
was injured but the aircraft sustained considerable damage to its
nose, oil cooler and exterior radio aerials. The pitot head was also
wiped out, and with it the airspeed indicator. The aircraft also lost
the use of an engine.

@@

On the night of Dec. 27, 1953, Squadron Leader E.L. Hare was homeward
bound from Japan in North Star 17516, bound for Shemya. The weather
was bad and the ceiling minimal. Visibility was a half mile with a
savage crosswind blowing at a 90-degree angle. The runway surface was
covered with wet snow; braking conditions were poor. On the third
landing attempt, everything came together in the worst possible
way. The North Star was blown off the runway into a gully. Everyone
survived, but 17516 was a total “write-off”–the only RCAF aircraft
lost in the course of Operation Hawk.

Canadian military policy with respect to honours and awards was very
restrictive from 1946 onwards, but rules were relaxed during the
Korean War. Personnel involved in Hawk were awarded one Officer of the
Order of the British Empire (OBE), one Member of the Order of the
British Empire (MBE), four Air Force Crosses, two Air Force Medals and
13 Queen’s Commendations. Most of these were to aircrew personnel, but
Sqdn. Ldr. W.H. Lord was granted an MBE for his work in establishing
and supervising the maintenance detachments that serviced the North
Stars wherever they alighted along the 11,000 mile route.

>>>>

@@

\textit{Photos courtesy of Library Archives Canada and 426 Thunderbird
  Sqn, RCAF.}\\

In 1947, while Prime Minister Mackenzie King was in London, his
Minister of Foreign Affairs (and chosen successor), Louis St-Laurent,
consented to Canadian participation on the United Nations Temporary
Commission of Korea (UNTOK). On his return to Canada, King furiously
confronted St-Laurent. He wanted no part---even at the fringes of
Korean affairs, which he considered dangerous.

@@

Mackenzie King died on July 22, 1950. As he lay in state in the
Parliament Buildings, six North Star aircraft of No. 426 Squadron flew
over, saluting his passing as they were en route to Tacoma,
Washington, to commence a Royal Canadian Air Force transoceanic
airlift in support of United Nations forces fighting in Korea.

It is not known if the former prime minister was spinning in his
coffin.

\begin{figure*}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{op-hawk-one.jpg}
   %caption of the figure 
   %\caption*{\small \em Jim Riddoch.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:op-hawk-one.eps}
\end{figure*}

@@

The North Star was basically a Douglas DC-4, but with Rolls-Royce
Merlin in-line engines instead of Pratt-and-Whitney radials. It was a
dependable aircraft, but generations of passengers and crew would
remember the tremendous noise and vibration in flight. The full story
of this is related by Larry Milberry in his 1982 book The Canadair
North Star.

@@

While the RCAF did not participate in the July 1948 to May 1949 Berlin
Airlift, a few individual aircrew did while on exchange duties with
the Royal Air Force and United States Air Force. As it happened, the
airlift seriously degraded the USAF Military Air Transport Service
(MATS) after flour and coal dust penetrated the seams of every
participating aircraft. When it was next marshalled to meet the Korean
crisis, MATS needed all the help it could get---and that was forthcoming
when the Canadian government committed No. 426 Sqdn.---its only
long-range transport unit---to the Korean campaign.

In recounting the RCAF's transpacific operational story, one must pay
tribute to Laurence Motiuk, author of two monumental volumes, namely
Thunderbirds at War: Diary of a Bomber Squadron, and Thunderbirds for
Peace: Diary of a Transport Squadron, which set out the history of
No. 426 Sqdn. The latter work was particularly difficult to write,
given that in the 1950s, RCAF units often submitted incomplete
historical narratives and in some instances neglected to prepare any
at all.

Motiuk's account of the postwar squadron was drawn from numerous
unofficial records, including log books as well as the sketchy unit
reports.

@@

Deployment to McChord Air Force Base, Tacoma, Washington, began on the
25th. Six North Stars, 12 crews and 185 ground personnel were
dispatched. Two-thirds of the aircrew were veterans of the Second
World War; the remainder were men who had joined the RCAF after
1947. Three of the aircraft left Tacoma on the 27th, arriving at
Haneda in Tokyo two days later. Operation Hawk was under way.

\begin{figure*}[htbp]
   \vspace{2em}
   \centering
   %name of the graphic, without the path AND in EPS format:
   \includegraphics[scale=0.5]{op-hawk-two.jpg}
   %caption of the figure 
   %\caption*{\small \em Jim Riddoch.}
   %label of the figure, which has to correspond to \ref{}:
   \label{fig:op-hawk-two.eps}
\end{figure*}

@@

The North Stars followed a challenging route. From McChord to
Elmendorf Air Force Base in Alaska was 1,490 miles. This was followed
by a 1,537-mile run to Shemya Air Force Base in the Aleutian
Islands. Next came the 2,104 mile flight to Haneda in Tokyo, with the
aircraft making landfall at Matsushima (Honshu, Japan).

@@

Between July 28 and Dec. 31, 1950, the squadron dispatched 123
aircraft across the Pacific. In the whole of 1951 it was 193 missions,
but in 1952 it dropped to 133. Not only were the Thunderbirds
returning to more traditional duties in Canada, they were also
assuming the job of providing transport, including personnel
rotation, to Canada's growing land and air presence in Europe. In
1953---the last full year of Hawk---it was 85 missions.

The final Hawk mission departed McChord on May 31, 1954, and ended at
the squadron's home base at Dorval on June 9, 1954. As it wound up, an
RCAF dispatch reported that No. 426 Sqdn. had flown 599 round
trips---four of them directly to Korea itself---logged 34,000 flying
hours, carried 13,000 personnel, and airlifted 3,500 tons of
freight. The report erred on two points. Six flights (not four) had
gone directly to Korea. The other discrepancy was that there were
only 584 missions; for some unknown reason the numbered listing of
flights had jumped from "439" on Nov. 26, 1952, to "455" on Nov. 28,
1952; the inaccurate 1954 press release continues to be quoted, ad
nauseam, to this day.

@@

On April 19, 1951, a North Star piloted by Flight Lieutenant J.A. Watt
was on a route-training flight to an unfamiliar field---Ashiya---located
between Osaka and Kobe, Japan. He was cleared by Ashiya tower to
descend from 4,000 to 3,000 feet. At 3,400 feet the aircraft hit trees
atop a hill that was not even indicated on the airfield map. No one
was injured but the aircraft sustained considerable damage to its
nose, oil cooler and exterior radio aerials. The pitot head was also
wiped out, and with it the airspeed indicator. The aircraft also lost
the use of an engine.

@@

On the night of Dec. 27, 1953, Squadron Leader E.L. Hare was homeward
bound from Japan in North Star 17516, bound for Shemya. The weather
was bad and the ceiling minimal. Visibility was a half mile with a
savage crosswind blowing at a 90-degree angle. The runway surface was
covered with wet snow; braking conditions were poor. On the third
landing attempt, everything came together in the worst possible
way. The North Star was blown off the runway into a gully. Everyone
survived, but 17516 was a total "write-off"---the only RCAF aircraft
lost in the course of Operation Hawk.

Canadian military policy with respect to honours and awards was very
restrictive from 1946 onwards, but rules were relaxed during the
Korean War. Personnel involved in Hawk were awarded one Officer of the
Order of the British Empire (OBE), one Member of the Order of the
British Empire (MBE), four Air Force Crosses, two Air Force Medals and
13 Queen's Commendations. Most of these were to aircrew personnel, but
Sqdn. Ldr. W.H. Lord was granted an MBE for his work in establishing
and supervising the maintenance detachments that serviced the North
Stars wherever they alighted along the 11,000 mile route.

### 2016-01-16 pnsac/newsletter/article_Feature2.tex
Adding files for April 2015 newsletter.

\title{The Whispering Giant}
\author{Dave Lambeth}

@@

Twinkle, twinkle, old North Star,\\
With wrinkled skin and creaky
spar,\\
For years you’ve always been on
sked,\\
As many of our force you’ve sped\\
to distant lands beyond the skies,\\
you’ve never failed to see the rise\\
of distant suns or moonrise glare.\\
You’ve carried us with tender care\\
-back you came to take us home,\\
from old Japan or France or Rome\\
from England, Goose, and Kef,\\
We cannot hear you.\\
(‘cause we’re deaf)\\

>>>>

\title{Volunteers Needed} \author{Richard Lodge}

@@

The Project North Star Association has two types of volunteers. The
first are known as Restoration volunteers and they are the group who
get their hands dirty and physically work on the restoration of the
aeroplane. The other group are known as Association volunteers and
they work in other very important areas of our Association.

At present we are not looking for any more Restoration volunteers but
we do have a need for some Association volunteers. In particular we
are looking for a Social Events Coordinator and a Merchandise Manager
and volunteers who would work with those officers.  These volunteer
posts have become open because the existing volunteers have held the
positions for some years and in each case have decided that it is time
to pursue other interests within the Association.

Below are brief details of the requirements for each volunteer
position. By volunteering for one of these functions, a volunteer will
become more involved with Association activities and will enjoy the
comradeship and rewards of working with a very interesting group of
people who have many diverse life experiences.

The Social Events Coordinator is the person responsible within the
Association for overseeing and organizing events for
members. Typically, he or she would, with the assistance of others,
organize the Members Meetings which are mostly held every three months
and such other events as he or she might agree to run as approved by
the Board of Directors.

The Merchandise Manager is responsible for managing the Association's
stock of merchandise for sale, setting the ordering levels as agreed
with the Board of Directors, placing orders for new stock and ensuring
that sales and inventory levels are properly reported to the
Treasurer. Other volunteers assist the Merchandise Manager in the
work.

If any of our members are interested in taking over one of these
positions or working with these officers or know of a non-member who
would be interested, please would he or she contact me by email at
\email{info@projectnorthstar.ca} or by phone on (613) 612-4920.

### 2016-01-16 pnsac/newsletter/article_PresReport.tex
Adding files for April 2015 newsletter.

During the past three months we have moved steadily forward with the
restoration of the North Star.  Our progress has been somewhat held up
by lack of available volunteers.

Our Association has two distinct and very important sides.  First and
foremost we are a group whose interest is in the North Star and the
restoration and preservation of the plane at the Museum.  Secondly,
over the years, we have developed a strong social side to the
Association where volunteers and other members can join in interesting
events centered around aviation. Elsewhere in this issue of the
Chronicle you will find a description by Bill Tate our Vice President,
of the very successful trip on October 4 to the Navcan Air Control
Centre and to the Bombardier factory in Montr\'{e}al.

After 10 years of restoration work on the North Star we recognize that
we must review our progress and develop a plan for the rest of the
work on the plane.  There will be three aspects to this review.  Our
first task will be to identify the major sections of the project and
develop ways of actually carrying out the restoration.  The
undercarriage of the plane is a particularly complicated challenge.
We then need to prepare a plan with projected timelines for completing
each part of the work.  Our third task will be to estimate costs and
to develop ideas for financing it.  During this process we must also
ensure that we will have sufficient qualified volunteers to get it
done.

Before the end of November we are hoping to have meetings with the
senior management of the Museum to update planning of all the tasks
above.  The end result of this should be a documented project plan and
a new Memorandum of Understanding between the Canada Aviation and
Space Museum and the Project North Star Association of Canada.

At times our progress on the restoration of the North Star may appear
to Association members to be somewhat slow.  Unlike other
organizations in Canada such as Vintage Wings in Gatineau or the
Canada Warplane Heritage Museum in Hamilton, PNSAC cannot make
decisions about the future of the North Star on its own.  As an
Association working with a Federal Government institution we are often
affected by decisions not related to the work we do.  This can at
times provide interesting challenges.

In conclusion, I would say that the Directors are working hard with
the Museum management to move our project forward even though at times
it may appear to be very difficult.  I hope that I will be able to
report significant progress in the next issue of the Chronicle.  In
the meantime we have some interesting events in the planning stages on
the social side of the Association.  Stay tuned.

>>>>

It is a year since I last wrote Notes from the President for the NStar Chronicle. During this time there have been many changes within the Project North Star Association.

In my last Notes, I referred to the temporary closure of restoration work by PNSAC volunteers on 
the North Star. We effectively did not restart operations until the middle of last year which 
resulted in a considerable loss of momentum for the Association as a whole together with the 
resignation of some of our directors and officers.

Last November we held our Annual General Meeting, which will in future be held in the Fall rather 
than in the summer when many of our members are on vacation. At this AGM two new directors were 
elected to replace two out of the three directors who had resigned.

We were very happy to welcome Roger Button and Phil Chrysler as new directors. Roger has been our corporate secretary. Phil has been our merchandise manager for several years and has also acted as assistant treasurer to the Association.

Drew Hodge, our long time NStar Chronicle publisher and webmaster, has after ten years, decided 
that he would like to spend more time on his other interests and did not stand for re-election as a director. Drew will continue to publish the NStar Chronicle and we are very pleased to 
welcome Garth Pyper as our new webmaster.

Bill Tate has been our vice president and  highly regarded special events coordinator for several years but decided that it was time to hand over to someone else.

Bruce Grant, who has ably editor the NS Chronicle for several years, also decided not to continue as a director and officer. Roger has agreed to take over the editor's  job where I am expecting that his word smithing abilities as a lawyer will come in handy.

The table officers of the Association  are now:  myself as president; Bruce Gemmill ,our senior project manager, as vice president; Roger  as secretary; and Paul Labranche as treasurer.

While the changes have been taking place with the executive of the Association, work has continued with our main reason for being; the restoration of the North Star. Elsewhere in this issue you will find a progress report on the work on the aircraft which as always is being done to a very high standard.

The North Star restoration project, now in its twelve year, will shortly be entering a new phase. 
Work on the fourth engine will be complete in a few months and we are now reworking our work plan 
for the restoration as a whole, in conjunction with the conservation professionals at Museum. In 
a future issue of the NStar Chronicle we hope to provide members with the broad outline of the 
updated work plan, together with anticipated time scales and funding requirements. Since the Museum does not receive government funds for the North Star restoration, future major work on the aircraft will have to be financed by donations from members and other people interested in preserving a plane which contributed so much to Canada?s military transportation needs in the late 1940s through to the early 1960s.

During the summer we hope to have two major displays of the North Star for Association members and the general public. The first will be during the weekend of June 6/7 when we will be taking part in Doors Open Ottawa and the second will be our usual display on Canada Day. As always, Association members will be particularly welcome at these displays and will be able to see the results of the excellent work being done by the volunteers on the plane.

### 2016-01-16 pnsac/newsletter/article_ProgReport.tex
Adding files for April 2015 newsletter.

The last engine to undergo restoration was removed in July and work
immediately began in the engine shop to disassemble the engine frame
and move the engine to the engine stand for restoration.  All cowl
panels, pipes, hoses and accessories have been removed and stored
until space is available to restore these items.  The priority has
been to disassemble the engine, which was completed in October, and
restoration and reassembly of engine 4 is now well underway.  The
refurbished crankshaft has been installed in the engine block, and the
connecting rods and pistons cleaned and polished.  Work has now begun
on the more complex cylinder heads and valve train assemblies.

Normally, work would also be underway to restore the engine frame, but
due to a lack of space, this will be delayed until the engine is back
together and can be moved out of the engine shop.  Some cleaning and
paint stripping has been done, but further work on the frame will
likely have to wait until more space is available for full restoration
of this large item.  This work will likely delay the completion of Nr
4 engine until 2016.

\section{Cocpit, Crew Lounge, Galley,  and Forward Washroom}
\label{crewlounge}

@@

The baggage compartments are now fully restored, and the refurbished
battery elevators were installed just aft of the nose wheel.  Work has
now begun on the main cabin.  This started with removal of various
fittings along the starboard side of the cabin, then removing the wood
paneling below the windows, and then removing the wood cargo floor
panels to expose the metal sub floor.  Most of the side panels and
some floor panels are badly damaged and will need to be replaced.  The
fittings for patient litters and troop seats were also removed.
Finally, the starboard windows were removed.  Once all the fittings
were removed, work began on repairing corrosion damage along the
exposed ribs and sub floor.

@@

Several badly corroded sections of the starboard side of the cabin
have been repaired, and work continues with corrosion repairs and
other cleanup while the aircraft is inside.  Part of the forward cargo
door has been removed for restoration.  The paratroop door insert is
complete, and the remaining portion of the forward cargo door is being
stripped and repaired prior to painting.  It is likely this will have
to be re-installed on the aircraft before all painting and restoration
is complete, so as not to delay it’s move outside.

@@

The North Star was originally fitted with troop seats down both sides
of the main cabin.  These could be stowed against the wall to make
room for cargo, or unrolled and set up to carry military personnel.
The original seats were removed by the RCAF, so a decision was made to
make a new set, using funds donated by Project North Star members.
Seat fabric, steel and aluminum tubing was ordered, and the troop seat
sections were fabricated, based on available drawings.  All seat
sections have been completed, and have been stored until the main
cabin is completed.

>>>>

The last engine to undergo restoration was removed in July and work
immediately began in the engine shop to disassemble the engine frame
and move the engine to the engine stand for restoration.  All cowl
panels, pipes, hoses and accessories have been removed and stored
until space is available to restore these items.  The priority has
been to disassemble the engine, which was completed in October, and
restoration and reassembly of engine 4 is now well under way.  The
refurbished crankshaft has been installed in the engine block, and the
connecting rods and pistons cleaned and polished.  Work has now begun
on the more complex cylinder heads and valve train assemblies.

Normally, work would also be under way to restore the engine frame, but
due to a lack of space, this will be delayed until the engine is back
together and can be moved out of the engine shop.  Some cleaning and
paint stripping has been done, but further work on the frame will
likely have to wait until more space is available for full restoration
of this large item.  This work will likely delay the completion of Nr
4 engine until 2016.

\section{Cockpit, Crew Lounge, Galley,  and Forward Washroom}
\label{crewlounge}

@@

The baggage compartments are now fully restored, and the refurbished
battery elevators were installed just aft of the nose wheel.  Work has
now begun on the main cabin.  This started with removal of various
fittings along the starboard side of the cabin, then removing the wood
panelling below the windows, and then removing the wood cargo floor
panels to expose the metal sub floor.  Most of the side panels and
some floor panels are badly damaged and will need to be replaced.  The
fittings for patient litters and troop seats were also removed.
Finally, the starboard windows were removed.  Once all the fittings
were removed, work began on repairing corrosion damage along the
exposed ribs and sub floor.

@@

Several badly corroded sections of the starboard side of the cabin
have been repaired, and work continues with corrosion repairs and
other cleanup while the aircraft is inside.  Part of the forward cargo
door has been removed for restoration.  The paratroop door insert is
complete, and the remaining portion of the forward cargo door is being
stripped and repaired prior to painting.  It is likely this will have
to be re-installed on the aircraft before all painting and restoration
is complete, so as not to delay its move outside.

@@

The North Star was originally fitted with troop seats down both sides
of the main cabin.  These could be stowed against the wall to make
room for cargo, or unrolled and set up to carry military personnel.
The original seats were removed by the RCAF, so a decision was made to
make a new set, using funds donated by Project North Star members.
Seat fabric, steel and aluminium tubing was ordered, and the troop seat
sections were fabricated, based on available drawings.  All seat
sections have been completed, and have been stored until the main
cabin is completed.
